# Grouping and Summarizing Data
## Grouping and Summarizing Data

When working with data, a very common task is to look at descriptive statistics for various groups.

We can use <span class='func'>group_by</span> to make this straightforward.

```{r groupby, eval=TRUE, cache=TRUE}
scoringDat %>% 
  group_by(Pos) %>% 
  summarize(meanTrueShooting = mean(trueShooting, na.rm = TRUE))
```


# Your Turn
## Putting it all together
?state.x77

Using one pipe sequence 

1. convert state.x77 (a base R object) to a data frame (<span class='func'>data.frame</span> )
2. create a new variable called **Region** that is equal to state.region (<span class='func'>mutate</span> )
3. create a new variable called **State** that is equal to state.name
4. <span class='func'>filter</span> only if Population is greater than 1000 (thousands)
5. <span class='func'>select</span> **Region** and variables beginning with I
6. group by **Region**  (<span class='func'>group_by</span> )
7. <span class='func'>summarize</span> **Income**, **Illiteracy** or **Both**, using the mean function

## Example
```{r yourturnDataManipExample, eval=TRUE, cache=TRUE}
state.x77 %>% 
  data.frame %>% 
  mutate(Region = state.region,
         State = state.name) %>% 
  filter(Population > 1000) %>% 
  select(Region, starts_with('I')) %>% 
  group_by(Region) %>% 
  summarize(meanInc=mean(Income),
            meanIll=mean(Illiteracy))
```

