## Grouping and Summarizing Data

## Grouping and Summarizing Data

A very common task is to look at group-based statistics

We can use <span class='func'>group_by</span> to make this straightforward

```{r groupby, eval=TRUE, cache=TRUE}
scoringDat %>% 
  group_by(Pos) %>% 
  summarize(meanTrueShooting = mean(trueShooting, na.rm = TRUE))
```


## Your Turn
## Putting it all together

`?state.x77`

Using one pipe sequence 

- $1.$ convert <span class="objclass">state.x77</span> (a base R object) to a data frame (<span class='func'>data.frame</span> )
- $2.$ create a new variable called **Region** that is equal to <span class="objclass">state.region</span> (<span class='func'>mutate</span> )
- $3.$ create a new variable called **State** that is equal to <span class="objclass">state.name</span>
- $4.$ <span class='func'>filter</span> only if Population is greater than 1000 (thousands)
- $5.$ <span class='func'>select</span> **Region** and variables beginning with the letter **I**
- $6.$ group by **Region**  (<span class='func'>group_by</span> )
- $7.$ <span class='func'>summarize</span> **Income**, **Illiteracy** or **Both**, using the <span class="func">mean</span> function

## Example
```{r yourturnDataManipExample, eval=TRUE, cache=TRUE}
state.x77 %>% 
  data.frame %>% 
  mutate(Region = state.region,
         State = state.name) %>% 
  filter(Population > 1000) %>% 
  select(Region, starts_with('I')) %>% 
  group_by(Region) %>% 
  summarize(meanInc=mean(Income),
            meanIll=mean(Illiteracy))
```

