---
title: "Exploring Data in R"
author: "Michael Clark <br> Center for Statistical Consultation and Research <br> University of Michigan"
date: "Fall, 2015"
output: 
  ioslides_presentation: 
    smaller: yes
    widescreen: yes
    theme: cosmo
    css: slideStyles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Exploring Data in R 

## Outline 
- Newer approaches to data wrangling
    - Introduction to plyr dplyr and tidyr
    - Subsetting rows
    - Subsetting columns
    - Reshaping data
    - Generating new data
    - Grouping and summarizing

- Nothin's gonna stop the flow
    - Piping with the magrittr package

- Quick interactive visualizations
    - Using *yr packages, piping and visualization
    - ggvis
    - htmlwidgets
    
    
# A Starting Point

## A Starting Point

```{r}
packs = c('magrittr', 'rvest', 'dplyr', 'stringr' ,'leaflet', 'ggvis')
sapply(packs, library, character.only=T)


wikiURL = 'https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population'

test0 = wikiURL %>% 
  read_html('wikitable sortable') %>% 
  html_table(fill=T) %>% 
  extract2(4) %>% 
  sapply(function(x) repair_encoding(as.character(x), 'UTF-8')) %>%
  data.frame %>%  
  mutate(latlon = sapply(str_split(Location, '/'), function(x) x[3]), 
         latlon = str_extract_all(latlon, '[-|[0-9]]+\\.[0-9]+'), 
         lat = sapply(latlon, function(x) x[1]),
         lon = sapply(latlon, function(x) x[2]), 
         population2014 = as.numeric(str_replace_all(X2014.estimate, ',', '')),
         population2010 = as.numeric(str_replace_all(X2010.Census, ',', '')),
         popDiff  = population2014-population2010) %>% 
  select(-latlon, -Location) %>% 
  filter(as.numeric(as.character(X2014.rank)) <= 100) #%>%  make all one when done



str(test0)

library(ggplot2)
test0 %>%
  leaflet %>% 
  addTiles %>% 
  addCircleMarkers(lat=~lat, lng=~lon, radius=~log(population2014))
```

