## Nothin's gonna stop the flow

## More with pipes

Recap thus far:

<span class='emph'>%>%</span> : Passes the prior object to the function after the pipe

- x <span class='emph'>%>%</span> f same as f(x)

Example:

```{r `%>%`}
iris %>% head
head(iris)
```


## More with pipes

Other pipes are available with <span class="pack">magrittr</span> and other packages

<span class='emph'>%$%</span>: Exposes the names in the prior to the function after

- x <span class='emph'>%\$%</span> y(a, b)  same as y(x\$a, x\$b)

Example (no data argument needed for <span class="func">lm</span>):

```{r `%$%`}
iris %$% lm(Sepal.Length ~ Sepal.Width)
```


## More with pipes

<span class='emph'>%T>%</span> : Passes to the next ***and*** subsequent steps

- x <span class='emph'>%T>%</span> y <span class='emph'>%>%</span> z is the same as x <span class='emph'>%>%</span> y & x <span class='emph'>%>%</span> z

## More with pipes

Example:

```{r Tpipe, fig.width=3, fig.height=3, echo=-1}
# par(mai=c(.1,.1,.1,.1), pch=19, byt='n')
iris %>% select(Sepal.Length, Sepal.Width) %T>% plot %>% summary
```


## More with pipes

Unfortunately the T pipe does not allow for some things

- Works:
```{r Tprobs1, eval=FALSE}
iris %>% select(Sepal.Length, Sepal.Width) %T>% plot %>% summary
```

- Provides no summary:
```{r Tprobs2, eval=FALSE}
iris %>% select(Sepal.Length, Sepal.Width) %T>% summary %>% plot
```

- Somewhat limiting in my opinion.


## More with pipes

<span class='emph'>%<>%</span> : assigns to former object the operations that follow

- Example:

```{r `%<>%`}
x = c(1:3,9:7,10,4:6)
x %<>% .^2 %>% sort
x
```

