# Reshaping Data
## Reshaping Data

Depending upon your analytical or visualization needs, sometimes you need to reshape your data.

Reshaping can take many forms. You might need to reshape your data from wide format to long format.

Or, maybe you need to split or combine variables.

Either way, R has you covered.

## Wide to Long

We are going to use the <span class='pack'>tidyr</span> package to make this data go from wide to long.

The function of note is <span class='func'>gather</span>.

- **key** is the new variable name, a factor whose labels are the variable names of the wide format
- **value** is the name of the variable that will contain their values

```{r wide2long, eval=TRUE, cache=TRUE, echo=-c(1)}
library(tidyr)
bballLong = bball %>% 
  select(Player, Tm, FG., X3P, X2P.1, trueShooting, effectiveFG) %>%
  rename(fieldGoalPerc = FG., threePointPerc = X3P, twoPointPerc = X2P.1) %>% 
  mutate(threePointPerc = threePointPerc/100) %>% 
  gather(key = 'vitalInfo', value = 'value', -Tm, -Player) 

bballLong %>% head
```


## Long to wide

Going the reverse direction

```{r long2wide, cache=TRUE, eval=TRUE}
bballLong %>%
  spread(vitalInfo, value) %>% 
  head
```