# Selecting Variables
## Selecting Variables

There are times when you do not want to look at the entire dataset, but instead want to focus on a few key variables.

Although this is easily handled in base R (as shown earlier), it can often more clearly accomplished using select in <span class='pack'>dplyr</span>

The following lets us look at the data clearly, without having to create objects, use quotes etc.

```{r select1, cache=TRUE}
bball %>% 
  select(Player, Tm, Pos, MP, trueShooting, effectiveFG, PTS) %>% 
  summary
```


## Selecting Variables
That works great, but now we need to drop some of those variables to look at correlations.

```{r select2, cache=TRUE}
scoringDat = bball %>%                                               # creating object as we will come back to it later
  select(Player, Tm, Pos, MP, trueShooting, effectiveFG, PTS)

scoringDat %>% 
  select(-Player, -Tm, -Pos) %>% 
  cor(use='complete') %>% 
  round(2)
```


## Selecting Variables
Sometimes, we have a lot of variables to select. If they have a common naming scheme, this becomes very easy.

```{r select3, cache=TRUE}
bball %>% 
  select(Player, contains("3P"), ends_with("RB")) %>% 
  arrange(desc(TRB)) %>% 
  head
```