## Selecting Variables

## Selecting Variables

Often you do not need the entire data set

Easily handled in base R (as shown earlier)

It can be more clear to use <span class="func">select</span> in <span class='pack'>dplyr</span>

- Don't have to create objects, use quotes or $, etc.

```{r select1, cache=TRUE}
bball %>% 
  select(Player, Tm, Pos, MP, trueShooting, effectiveFG, PTS) %>% 
  summary
```


## Selecting Variables

What if we want to drop some variables?

- Can't calculate correlation on the string variables, so drop

```{r select2, cache=TRUE}
scoringDat = bball %>%                                               # creating object as we will come back to it later
  select(Player, Tm, Pos, MP, trueShooting, effectiveFG, PTS)

scoringDat %>% 
  select(-Player, -Tm, -Pos) %>% 
  cor(use='complete') %>% 
  round(2)
```


## Selecting Variables

Sometimes, we have a lot of variables to select

If they have a common naming scheme, this can be very easy

```{r select3, cache=TRUE}
bball %>% 
  select(Player, contains("3P"), ends_with("RB")) %>% 
  arrange(desc(TRB)) %>% 
  head
```